---
title: "Data Exploration and Modeling"
author: "Ada Lazuli"
date: '2022-07-06'
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DT)
library(rpart)
library(ggplot2)
library(party)
library(tidymodels)
library(rattle)
library("ggfortify")
set.seed(101011)
```


# Helper Functions

This section details the helper function that were used to reduce come of the repetition during exploratory data analysis.

```{r, out.width = '100%'}
# Create a histogram of the variable with annotations for the mean and outliers
generate_histogram <- function(df, col, title, x) {
  ggplot(df, aes(x = col)) + geom_histogram() + 
    geom_vline(aes(xintercept = mean(col)), color = "red") + 
    geom_vline(aes(xintercept = mean(col)  - 3 * sd(col)), color = "blue") + 
    geom_vline(aes(xintercept = mean(col)  +  3 * sd(col)), color = "blue") + 
    labs(title=title, x = x, y = "Occurrences")
}

generate_boxplot <- function(df, col, title, x) {
  ggplot(df, aes(x = col)) + geom_boxplot() + 
    labs(title=title, x = x, y = "Occurrences")
}

generate_class_histogram_facet <- function(df, col, title, x){
  ggplot(df, aes(x = col, fill = CLASS)) +
    geom_histogram() + coord_flip() +
    facet_grid(~ CLASS) +
    labs(title=title, x = x, y = "Occurrences")
}


generate_class_boxplot_facet <- function(df, col, title, x){
  ggplot(df, aes(x = col, fill = CLASS)) + 
    geom_boxplot() + coord_flip() + 
    facet_grid(~ CLASS) +
    labs(title=title, x = x)
}

generate_nsp_histogram_facet <- function(df, col, title, x){
  ggplot(df, aes(x = col, fill = NSP)) +
    geom_histogram() + coord_flip() +
    facet_grid(~ NSP) +
    labs(title=title, x = x, y = "Occurrences")
}


generate_nsp_boxplot_facet <- function(df, col, title, x){
  ggplot(df, aes(x = col, fill = NSP)) + 
    geom_boxplot() + coord_flip() + 
    facet_grid(~ NSP) +
    labs(title=title, x = x)
}



return_outliers <- function(data) {
  upper_limit <- mean(data) + 3 * sd(data)
  lower_limit <- mean(data) - 3 * sd(data)
  
  mask <- (data > upper_limit)  | (data < lower_limit)
  
  return (data[mask])
}

```

# Data Exploration

```{r, out.width = '100%'}
df <- read.csv("CTG.csv")
# convert the classes to factors
df$CLASS <- as.factor(df$CLASS)
df$NSP <- as.factor(df$NSP)

```

The data consists of `r nrow(df)` rows that detail the results of Cardiotocograms (CTGs)  and a classification by expert obstetricians (Marques de Sai et al, 2010). For each row, the data contains `r ncol(df)` variables, with the final two being the classifications by experts (Marques de Sai et al, 2010).

```{r, out.width = '100%'}
str(df)
datatable(df)
```


## LB - FHR baseline (beats per minute)

The __LB__ variable details the fetal heart rate baseline in terms of beats per minute (Marques de Sai et al, 2010).

### Overview

```{r warning=FALSE}
col_of_interest <- df$LB
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Fetal Heartbeat Baseline", "Heartbeat Baseline") 
generate_boxplot(df, col_of_interest,  "Histogram of Fetal Heartbeat Baseline", "Heartbeat Baseline") 
```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of LB by Morphologic Pattern (CLASS)", "Heartbeat Baseline")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of LB by Morphologic Pattern (CLASS)", "Heartbeat Baseline")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of LB by Fetal State Class", "Heartbeat Baseline")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of LB by Fetal State Class", "Heartbeat Baseline")
```


## AC - Accelerations per second

The __AC__ variable details the accelerations per second as recorded during the CTG. (Marques de Sai et al, 2010)

### Overview

```{r warning=FALSE}
col_of_interest <- df$AC
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Fetal Accelerations per Second ", "Accelerations per Second ") 
generate_boxplot(df, col_of_interest,  "Histogram of Fetal Accelerations per Second", "Accelerations per Second") 

```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of AC by Morphologic Pattern (CLASS)", "Accelerations per Second")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of AC by Morphologic Pattern (CLASS)", "Accelerations per Second")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of AC by Fetal State Class", "Accelerations per Second")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of AC by Fetal State Class", "Accelerations per Second")
```



## FM - Fetal movements per second

The __FM__ variable details the fetal movements per second (Marques de Sai et al, 2010).

### Overview

```{r warning=FALSE}
col_of_interest <- df$FM
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Fetal Movements per Second", "Fetal Movements per Second") 
generate_boxplot(df, col_of_interest,  "Histogram of Fetal Movements per Second", "Fetal Movements per Second") 

```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of FM by Morphologic Patternz (CLASS)", "Fetal Movements per Second")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of FM by Morphologic Pattern (CLASS)", "Fetal Movements per Second")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of FM by Fetal State Class", "Fetal Movements per Second")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of FM by Fetal State Class", "Fetal Movements per Second")
```



## UC - Uterine contractions per second

The __UC__ variable details the uterine contractions per second  (Marques de Sai et al, 2010).

### Overview

```{r warning=FALSE}
col_of_interest <- df$UC
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Uterine Contractions Per Second", "Uterine Contractions Per Second") 
generate_boxplot(df, col_of_interest,  "Histogram of Uterine Contractions Per Second", "Uterine Contractions Per Second") 

```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of UC by Morphologic Pattern (CLASS)", "Uterine Contractions Per Second")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of UC by Morphologic Pattern (CLASS)", "Uterine Contractions Per Second")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of UC by Fetal State Class", "Uterine Contractions Per Second")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of UC by Fetal State Class", "Uterine Contractions Per Second")
```



## DL - Light decelerations per second

The __DL__ variable details light deceleration per second as recorded during the CTG  (Marques de Sai et al, 2010).

### Overview

```{r warning=FALSE}
col_of_interest <- df$DL
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Decelerations per Second", "Decelerations per Second") 
generate_boxplot(df, col_of_interest,  "Histogram of Decelerations per Second", "Decelerations per Second") 

```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of DL by Morphologic Pattern (CLASS)", "Decelerations per Second")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of DL by Morphologic Pattern (CLASS)", "Decelerations per Second")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of DL by Fetal State Class", "Decelerations per Second")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of DL by Fetal State Class", "Decelerations per Second")
```

## DS - Severe decelerations per second

The __DS__ variable details the accelerations per second

### Overview

```{r warning=FALSE}
col_of_interest <- df$DS
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Severe decelerations per second", "Severe decelerations per second") 
generate_boxplot(df, col_of_interest,  "Histogram of Severe decelerations per second", "Severe decelerations per second") 

```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of DS by Morphologic Pattern (CLASS)", "Severe decelerations per second")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of DS by Morphologic Pattern (CLASS)", "Severe decelerations per second")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of DS by Fetal State Class", "Severe decelerations per second")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of DS by Fetal State Class", "Severe decelerations per second")
```


## DP - Prolongued Decelerations per Second

The __DP__ variable details the prolongued decelerations per second, as recorded during the CTG  (Marques de Sai et al, 2010).

### Overview

```{r warning=FALSE}
col_of_interest <- df$DP
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Prolongued Decelerations per Second", "Prolongued Decelerations per Second") 
generate_boxplot(df, col_of_interest,  "Histogram of Prolongued Decelerations per Second", "Prolongued Decelerations per Second") 

```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of DP by Morphologic Pattern (CLASS)", "Prolongued Decelerations per Second")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of DP by Morphologic Pattern (CLASS)", "Prolongued Decelerations per Second")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of DP by Fetal State Class", "Prolongued Decelerations per Second")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of DP by Fetal State Class", "Prolongued Decelerations per Second")
```



## ASTV - Percentage of Time with Abnormal Short Term Variability

The __ASTV__ variable details the percentage of time with abnormal short term variability (Marques de Sai et al, 2010).

### Overview

```{r warning=FALSE}
col_of_interest <- df$ASTV
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Percentage of Time with Abnormal Short Term Variability", "Percentage of Time with Abnormal Short Term Variability") 
generate_boxplot(df, col_of_interest,  "Histogram of Percentage of Time with Abnormal Short Term Variability", "Percentage of Time with Abnormal Short Term Variability") 

```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of ASTV by Morphologic Pattern (CLASS)", "Percentage of Time with Abnormal Short Term Variability")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of ASTV by Morphologic Pattern (CLASS)", "Percentage of Time with Abnormal Short Term Variability")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of ASTV by Fetal State Class", "Percentage of Time with Abnormal Short Term Variability")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of ASTV by Fetal State Class", "Percentage of Time with Abnormal Short Term Variability")
```



## MSTV - mean value of short term variability

The __MSTV__ variable details the mean value of short term variability  (Marques de Sai et al, 2010).

### Overview

```{r warning=FALSE}
col_of_interest <- df$MSTV
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Mean Short Term Variability", "Mean Short Term Variability") 
generate_boxplot(df, col_of_interest,  "Histogram of Mean Short Term Variability", "Mean Short Term Variability") 

```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of MSTV by Morphologic Pattern (CLASS)", "Mean Short Term Variability")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of MSTV by Morphologic Pattern (CLASS)", "Mean Short Term Variability")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of MSTV by Fetal State Class", "Mean Short Term Variability")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of MSTV by Fetal State Class", "Mean Short Term Variability")
```


## ALTV - percentage of time with abnormal long term variability

The __ALTV__ variable details the percentage of time with abnoramal long term variability  (Marques de Sai et al, 2010).

### Overview

```{r warning=FALSE}
col_of_interest <- df$ALTV
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Percentage of Time with Abnormal Long Term Variability", "Percentage of Time with Abnormal Long Term Variability") 
generate_boxplot(df, col_of_interest,  "Histogram of Percentage of Time with Abnormal Long Term Variability", "Percentage of Time with Abnormal Long Term Variability") 

```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of ALTV by Morphologic Pattern (CLASS)", "Percentage of Time with Abnormal Long Term Variability")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of ALTV by Morphologic Pattern (CLASS)", "Percentage of Time with Abnormal Long Term Variability")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of ALTV by Fetal State Class", "Percentage of Time with Abnormal Long Term Variability")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of ALTV by Fetal State Class", "Percentage of Time with Abnormal Long Term Variability")
```



## MLTV - mean value of long term variability


The __MLTV__ variable details the mean value of long term variability  (Marques de Sai et al, 2010).

### Overview

```{r warning=FALSE}
col_of_interest <- df$MLTV
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Mean  Long Term Variability", "Mean  Long Term Variability") 
generate_boxplot(df, col_of_interest,  "Histogram of Mean  Long Term Variability", "Mean  Long Term Variability") 

```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of MLTV by Morphologic Pattern (CLASS)", "Mean  Long Term Variability")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of MLTV by Morphologic Pattern (CLASS)", "Mean  Long Term Variability")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of MLTV by Fetal State Class", "Mean  Long Term Variability")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of MLTV by Fetal State Class", "Mean  Long Term Variability")
```



## Width - width of FHR histogram

The __Width__ variable details the width of the CTG heart rate histogram  (Marques de Sai et al, 2010).

### Overview

```{r warning=FALSE}
col_of_interest <- df$Width
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Width of FHR Histogram", "Width of FHR Histogram") 
generate_boxplot(df, col_of_interest,  "Histogram of Width of FHR Histogram", "Width of FHR Histogram") 

```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of Width by Morphologic Pattern (CLASS)", "Width of FHR Histogram")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of Width by Morphologic Pattern (CLASS)", "Width of FHR Histogram")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of Width by Fetal State Class", "Width of FHR Histogram")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of Width by Fetal State Class", "Width of FHR Histogram")
```



## Min - minimum of FHR histogram

The __Min__ variable details the minimum in the CTG heart rate histogram  (Marques de Sai et al, 2010).

### Overview

```{r warning=FALSE}
col_of_interest <- df$Min
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Min of FHR Histogram", "Min of FHR Histogram") 
generate_boxplot(df, col_of_interest,  "Histogram of Min of FHR Histogram", "Min of FHR Histogram") 

```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of Min by Morphologic Pattern (CLASS)", "Min of FHR Histogram")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of Min by Morphologic Pattern (CLASS)", "Min of FHR Histogram")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of Min by Fetal State Class", "Min of FHR Histogram")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of Min by Fetal State Class", "Min of FHR Histogram")
```



## Max - Maximum of FHR histogram

The __Max__ variable details the maximum in the CTG heart rate histogram  (Marques de Sai et al, 2010).

### Overview

```{r warning=FALSE}
col_of_interest <- df$Max
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Max of FHR Histogram", "Max of FHR Histogram") 
generate_boxplot(df, col_of_interest,  "Histogram of Max of FHR Histogram", "Max of FHR Histogram") 

```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of Max by Morphologic Pattern (CLASS)", "Max of FHR Histogram")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of Max by Morphologic Pattern (CLASS)", "Max of FHR Histogram")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of Max by Fetal State Class", "Max of FHR Histogram")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of Max by Fetal State Class", "Max of FHR Histogram")
```



## Nmax - # of histogram peaks

The __Nmax__ variable details the number of peaks in the CTG heart rate histogram  (Marques de Sai et al, 2010).

### Overview

```{r warning=FALSE}
col_of_interest <- df$Nmax
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Number of Histogram Peaks", "Number of Histogram Peaks") 
generate_boxplot(df, col_of_interest,  "Histogram of Number of Histogram Peaks", "Number of Histogram Peaks") 

```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of Nmax by Morphologic Pattern (CLASS)", "Number of Histogram Peaks")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of Nmax by Morphologic Pattern (CLASS)", "Number of Histogram Peaks")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of Nmax by Fetal State Class", "Number of Histogram Peaks")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of Nmax by Fetal State Class", "Number of Histogram Peaks")
```




## Nzeros - # of histogram zeros

The __Nzeros__ variable details the number of zeros in the CTG heart rate histogram  (Marques de Sai et al, 2010).

### Overview

```{r warning=FALSE}
col_of_interest <- df$Nzeros
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Number of Histogram Zeros", "Number of Histogram Zeros") 
generate_boxplot(df, col_of_interest,  "Histogram of Number of Histogram Zeros", "Number of Histogram Zeros") 

```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of Nzeros by Morphologic Pattern (CLASS)", "Number of Histogram Zeros")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of Nzeros by Morphologic Pattern (CLASS)", "Number of Histogram Zeros")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of Nzeros by Fetal State Class", "Number of Histogram Zeros")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of Nzeros by Fetal State Class", "Number of Histogram Zeros")
```



## Mode - histogram mode

The __Mode __ variable details the mode of the CTG heart rate histogram  (Marques de Sai et al, 2010).

### Overview

```{r warning=FALSE}
col_of_interest <- df$Mode 
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Mode of FHR Histogram", "Mode of FHR Histogram") 
generate_boxplot(df, col_of_interest,  "Histogram of Mode of FHR Histogram", "Mode of FHR Histogram") 

```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of Mode  by Morphologic Pattern (CLASS)", "Mode of FHR Histogram")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of Mode  by Morphologic Pattern (CLASS)", "Mode of FHR Histogram")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of Mode  by Fetal State Class", "Mode of FHR Histogram")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of Mode  by Fetal State Class", "Mode of FHR Histogram")
```



## Mean - histogram mean

The __Mean __ variable details the mean of the CTG heart rate histogram  (Marques de Sai et al, 2010).

### Overview

```{r warning=FALSE}
col_of_interest <- df$Mean 
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Mean of FHR Histogram", "Mean of FHR Histogram") 
generate_boxplot(df, col_of_interest,  "Histogram of Mean of FHR Histogram", "Mean of FHR Histogram") 

```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of Mean  by Morphologic Pattern (CLASS)", "Mean of FHR Histogram")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of Mean  by Morphologic Pattern (CLASS)", "Mean of FHR Histogram")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of Mean  by Fetal State Class", "Mean of FHR Histogram")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of Mean  by Fetal State Class", "Mean of FHR Histogram")
```



## Median - histogram median

The __Median __ variable details the median of the CTG heart rate histogram  (Marques de Sai et al, 2010).

### Overview

```{r warning=FALSE}
col_of_interest <- df$Median 
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Median of FHR Histogram", "Median of FHR Histogram") 
generate_boxplot(df, col_of_interest,  "Histogram of Median of FHR Histogram", "Median of FHR Histogram") 

```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of Median  by Morphologic Pattern (CLASS)", "Median of FHR Histogram")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of Median  by Morphologic Pattern (CLASS)", "Median of FHR Histogram")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of Median  by Fetal State Class", "Median of FHR Histogram")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of Median  by Fetal State Class", "Median of FHR Histogram")
```




## Variance - histogram variance

The __Variance __ variable details the variance of the CTG heart rate histogram  (Marques de Sai et al, 2010).

### Overview

```{r warning=FALSE}
col_of_interest <- df$Variance 
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Variance of FHR Histogram", "Variance of FHR Histogram") 
generate_boxplot(df, col_of_interest,  "Histogram of Variance of FHR Histogram", "Variance of FHR Histogram") 

```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of Variance  by Morphologic Pattern (CLASS)", "Variance of FHR Histogram")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of Variance  by Morphologic Pattern (CLASS)", "Variance of FHR Histogram")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of Variance  by Fetal State Class", "Variance of FHR Histogram")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of Variance  by Fetal State Class", "Variance of FHR Histogram")
```



## Tendency - histogram tendency

The __Tendency __ variable details the tendency of the CTG heart rate histogram  (Marques de Sai et al, 2010).

### Overview

```{r warning=FALSE}
col_of_interest <- df$Tendency 
summary(col_of_interest)
generate_histogram(df, col_of_interest,  "Histogram of Tendency of FHR Histogram", "Tendency of FHR Histogram") 
generate_boxplot(df, col_of_interest,  "Histogram of Tendency of FHR Histogram", "Tendency of FHR Histogram") 

```

There were `r length(return_outliers(col_of_interest))` outliers for this column.


### Facet By Morphologic Pattern (CLASS)

```{r, out.width = '100%'}
generate_class_histogram_facet(df, col_of_interest, "Hisogram of Tendency  by Morphologic Pattern (CLASS)", "Tendency of FHR Histogram")
generate_class_boxplot_facet(df, col_of_interest, "Boxplot of Tendency  by Morphologic Pattern (CLASS)", "Tendency of FHR Histogram")
```


### Facet By Fetal State Class (NSP)

```{r, out.width = '100%'}
generate_nsp_histogram_facet(df, col_of_interest, "Histogram of Tendency  by Fetal State Class", "Tendency of FHR Histogram")
generate_nsp_boxplot_facet(df, col_of_interest, "Boxplot of Tendency  by Fetal State Class", "Tendency of FHR Histogram")
```


## CLASS - FHR pattern class code (1 to 10)


```{r, out.width = '100%'}
table(df$CLASS)
```


## NSP - fetal state class code (N=normal; S=suspect; P=pathologic)

```{r, out.width = '100%'}
table(df$NSP)
```

# Data Loading

Load the data for modeling. During exploratory analysis, it was discovered that the classifications for __NSP__ are not balanced. The `initial_split` function from `rsample` in `tidymodels` is used to split the data into training and test sets, while preserving the distribution of the __NSP__ variable (Silge).

```{r, out.width = '100%'}
df <- read.csv("CTG.csv")
# convert the classes to factors
df$CLASS <- NULL
df$NSP <- as.factor(df$NSP)
data_partitioned <- initial_split(df, prop = 0.75, strata = NSP)
train <- training(data_partitioned)
test <-  testing(data_partitioned)
```

# Tree Creation

## First Attempt

### Model Definition

The first attempt involved using all of the available features and not specifying any limitations to the model. The tree is created using the `parsnip` package in `tidymodels`, with the _rpart_ engine and set for classification (Kuhn). The tree was fit on the data, using all of the available columns

```{r, out.width = '100%'}
tree_template <- decision_tree() %>% set_engine("rpart") %>% set_mode("classification")
tree_model <- tree_template %>% fit(formula = NSP ~ ., data =  train)
fancyRpartPlot(tree_model$fit, caption = "First Decision Tree Attempt")
```

### Model Results

To assess the performance of the tree, the accuracy, confusion matrix, ROC Curve, and AUC are all captured (Han et al, 2011, p. 49).

```{r, out.width = '100%'}
# For the confusion Matrix
predictions <- predict(tree_model, test) %>% mutate(true = test$NSP)
# For the plot of the ROC Curve
predictions_prob <- predict(tree_model, test, type = "prob") %>% bind_cols(test)
accuracy(data = predictions, estimate = .pred_class, truth = true)
conf_mat(data = predictions, estimate = .pred_class, truth = true)
autoplot(roc_curve(data = predictions_prob, estimate = c(.pred_1, .pred_2, .pred_3), truth = NSP)) + ggtitle("Multi Class ROC Curves For Model 1")
print(roc_auc(data = predictions_prob, estimate = c(.pred_1, .pred_2, .pred_3), truth = NSP))
```

__Note__: The article written by Brendan Cullen (2021) [here](https://bcullen.rbind.io/post/2020-06-02-tidymodels-decision-tree-learning-in-r/) helped a bit with using features available in the collection packages found in tidymodels. 

## Second Attempt

### Optimization

The second attempt is to use a grid search to find the optimal combination of _min_n, tree depth, and cost complexity_ for the model using `tune_grid` from the `tidymodels` set of packages (Kuhn). 

```{r, out.width = '100%'}
tune_specification <- decision_tree(tree_depth = tune(), min_n = tune(), cost_complexity = tune()) %>% set_mode("classification") %>% set_engine("rpart")

grid_search <- grid_regular(parameters(tune_specification), levels = 10)

tuned <- tune_grid(tune_specification, NSP ~ ., resample = vfold_cv(train, v = 3), grid = grid_search, metrics = metric_set(accuracy))

autoplot(tuned)
```

### Using Best Parameters

Following the grid search, the best performing set of parameters were saved and used to create a second model.

```{r, out.width = '100%'}
optimal_parameters <- select_best(tuned)
print(optimal_parameters)
optimal_tree_specification <- finalize_model(tune_specification, optimal_parameters)

optimal_model <- fit(optimal_tree_specification,
                   NSP ~ .,
                   train)

fancyRpartPlot(optimal_model$fit, caption = "Final Decision Tree Attempt")
```

### Model Performance

To assess the performance of the tree, the accuracy, confusion matrix, ROC Curve, and AUC are all captured (Han et al, 2011, p. 49).


```{r, out.width='100%'}
predictions <- predict(optimal_model, test) %>% mutate(true = test$NSP)
predictions_prob <- predict(optimal_model, test, type = "prob") %>% bind_cols(test)
accuracy(data = predictions, estimate = .pred_class, truth = true)
conf_mat(data = predictions, estimate = .pred_class, truth = true)
autoplot(roc_curve(data = predictions_prob, estimate = c(.pred_1, .pred_2, .pred_3), truth = NSP)) + ggtitle("Multi Class ROC Curves For Final Model")
print(roc_auc(data = predictions_prob, estimate = c(.pred_1, .pred_2, .pred_3), truth = NSP))
```

# References 

Han, Kamber, & Pei. (2011). Chapter 8. Classification: Basic Concepts. Elsevier Science. 
Kuhn, M. (n.d.) _Model Tuning Via Grid Search_ Retrieved from: https://tune.tidymodels.org/reference/tune_grid.html
Kuhn, M. (n.d.) _Decision Trees_. Retrieved from: https://parsnip.tidymodels.org/reference/decision_tree.html
Marques de Saj, J., Bernardes, J,, & Aryres de Campos, D. (2010). _Cardiotocography Data Set [CSV]_. Retrieved from: https://archive.ics.uci.edu/ml/datasets/Cardiotocography#
Silge, J. (n.d). _Simple Training/Test Set Splitting_. Retrieved from: https://rsample.tidymodels.org/reference/initial_split.html
